---
output: github_document
bibliography: vignettes/References.bib
---

<!-- *********** -->
<!-- Note: README.md is generated from README.Rmd.   -->
<!-- Be sure to edit README.Rmd and generate the README.md file by Cmd/Ctl-shift-K -->
<!-- *********** -->

[![CRAN status](https://www.r-pkg.org/badges/version/matsindf)](https://cran.r-project.org/package=matsindf)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis-CI Build Status](https://travis-ci.org/MatthewHeun/matsindf.svg?branch=master)](https://travis-ci.org/MatthewHeun/matsindf)
[![Coverage status](https://codecov.io/gh/MatthewHeun/matsindf/branch/master/graph/badge.svg)](https://codecov.io/github/MatthewHeun/matsindf?branch=master)


# matsindf

The `R` package `matsindf` provides 
ways to interact with matrices in data frames. 


## Installation

You can install `matsindf` from CRAN with:

```{r CRAN-installation, eval = FALSE}
install.packages("matsindf")
```

You can install a recent development version of `matsindf` from github with:
```{r gh-installation, eval = FALSE}
# install devtools if not already installed
# install.packages("devtools")
devtools::install_github("MatthewHeun/matsindf")
# To build vignettes locally, use
devtools::install_github("MatthewHeun/matsindf", build_vignettes = TRUE)
```


## History

The functions in this package were used in 
[Heun et al. [-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).


## More Information

Find more information, including vignettes and function documentation, at 
<https://MatthewHeun.github.io/matsindf/>.


## References
